Permutation Pattern matching in\\ $(213,231)$-avoiding permutations

### 2016-06-22 article.tex
modif des reviews 2

@@

%\DeclareMathOperator{\run}{run}
\DeclareMathOperator{\block}{block}
\DeclareMathOperator{\LMEi}{LMEi}

\DeclareMathOperator{\factor}{factor}
%\newcommand{\factor}[2]{\factora_{{#1}}({#2})}

@@

%Pattern Marching
\DeclareMathOperator{\PMa}{PM}
\newcommand{\PM}[6]{\PMa_{{#1}}^{{#2},{#3},{#4}}({#5},{#6})}

@@

\title{Pattern matching in\\ $(213,231)$-avoiding permutations}
\author{Both Emerite NEOU, Romeo RIZZI}
\date{}

\author{%
	Both Emerite Neou\thanks{On a Co-tutelle  Agreement with the Department of Mathematics of the University of Trento}\inst{1}
  Romeo Rizzi\inst{2} \and
  St\'ephane Vialette\inst{1}
}% end author
\institute{%
	Universit\'e Paris-Est, LIGM (UMR 8049), CNRS, UPEM, ESIEE Paris, ENPC,
	F-77454, Marne-la-Vallée, France\\
  \email{\{neou,vialette\}@univ-mlv.fr}
  \and
  Department of Computer Science,
  Università degli Studi di Verona, Italy \\
  \email{romeo.rizzi@univr.it}
}% end institute

@@

\begin{abstract}
Given permutations $\sigma \in S_k$ and $\pi \in S_n$ with $k<n$, the
\emph{pattern matching} problem is to decide whether $\sigma$ occurs
$\pi$ as an order-isomorphic subsequence.
We give a linear-time algorithm in case both $\pi$ and $\sigma$ avoid
the two size-$3$ permutations $213$ and $231$.
For the special case where only $\sigma$ avoids $213$ and $231$, we present a
$O(max(kn^2,n^2\log(\log(n)))$ time algorithm. 
We extend our research to bivincular patterns that avoid $213$ and $231$ and present a $O(kn^4)$ time algorithm.
Finally we look at the related problem of the longest subsequence which avoids $213$ and $231$.  
\end{abstract}

@@

A permutation $\sigma$ is said to occur in another permutation $\pi$,
denoted $\sigma \preceq \pi$,
if there exists a subsequence of elements of $\pi$ that has the same relative
order as $\sigma$.
Otherwise, $\pi$ is said to \emph{avoid} the permutation $\sigma$.
For example a permutation has an occurrence of the pattern $123$ (resp. $321$) if it has
an increasing (resp. decreasing) subsequence of length $3$.
Similarly,
$6152347$ has an occurrence of $213$ but not $231$.
During the last decade, the study of the pattern matching for permutations has
become a very active area of research \cite{Kitaev:book:2011}. and
a whole annual conference (\textsc{Permutation Patterns}) focuses
on pattern in permutation.

We consider here the so-called \emph{pattern matching} problem
(also sometimes referred to as the \emph{pattern involvement problem}):
Given two permutations $\sigma$ and $\pi$, this problem is to decide whether
$\sigma \preceq \pi$ (the problem is attributed to Wilf in \cite{Bose:Buss:Lubiw:1998}).
The pattern matching for permutations is known to be \NPhard~\cite{Bose:Buss:Lubiw:1998}.
It is, however, polynomial-time solvable by brute-force enumeration
if $\sigma$ has bounded size.
Improvements to this algorithm were presented in
\cite{Albert:Aldred:Atkinson:Holton:ISAAC:2001} and
\cite{Ahal:Rabinovich:2008},
the latter describing a $O(|\pi|^{0.47|\sigma|+o(|\sigma|)})$ time algorithm.
Bruner and Lackner \cite{DBLP:journals/corr/abs-1204-5224}
gave a fixed-parameter algorithm solving the pattern matching for permutations problem with
an exponential worst-case runtime of $O(1.52^{n}.n.k)$,
(This is an improvement upon the $O(2^{|\pi|})$ runtime required by
brute-force search without imposing restrictions on $\sigma$ and $\pi$.)
A recent major step was taken by Marx and Guillemot
\cite{Guillemot:Marx:SODA:2014}.
They showed that
the pattern matching for permutations problem is fixed-parameter tractable (FPT) for
parameter $|\sigma|$.

A few particular cases of the pattern matching for permutations problem have been attacked successfully.
The case of increasing patterns is solvable in
$O(|\pi| \log \log |\sigma|)$ time in the RAM model \cite{Crochemore:Porat:2010},
improving the previous 30-year bound of $O(|\pi| \log |\sigma|)$.
% (The algorithm also improves on the previous
% $O(|\pi| \log \log |\pi|)$ bound.)
Furthermore, the patterns $132$, $213$, $231$, $312$ can all be handled in linear-time
by stack sorting algorithms.
Any pattern of length $4$ can be detected in $O(|\pi| \log |\pi|)$ time
\cite{Albert:Aldred:Atkinson:Holton:ISAAC:2001}.
Algorithmic issues for $321$-avoiding patterns matching for permutations have been investigated in
\cite{Guillemot:Vialette:ISAAC:2009} and more recently in \cite{2015arXiv151006051A}.
The pattern matching for permutation problem is also solvable in
polynomial-time for separable patterns \cite{Ibarra:1997,Bose:Buss:Lubiw:1998}
(see also \cite{Bouvel:Rossin:Vialette:CPM:2007} for Longest common subsequence problem and related one on separable permutations).
Separable permutations are permutations that do not have an occurrence of
$2413$ nor $3142$, and they are enumerated by the Schr{\"o}der numbers.
%	(sequence A006318 in OEIS).
%	To the best of our knowledge,
%	separable permutations first arose in the work of
%	Avis and Newborn~\cite{Avis:Newborn:1981},
%	who showed that they are precisely the permutations which can be sorted by an
%	arbitrary number of pop-stacks in series,
%	where a pop-stack is a restricted form of stack in which any pop operation
%	pops all items at once.
(Notice that the separable permutations include as a special case the
stack-sortable permutations, which avoid the pattern $231$.)

There exist many generalisations of patterns that are worth considering
in the context of algorithmic issues in pattern matching
(see \cite{Kitaev:book:2011} for an up-to-date survey).
\emph{Vincular patterns}, also called
\emph{generalized patterns},
resemble (classical) patterns with the additional constraint that some of the elements in
a occurrence must be consecutive in postitions.
Of particular importance in our context,
Bruner and Lackner \cite{DBLP:journals/corr/abs-1204-5224}
proved that deciding whether a vincular pattern
$\sigma$ of length $k$ can occur in a longer permutation
$\pi$ is $W[1]$-complete for
parameter $k$;
for an up-to-date survey of the $W[1]$ class and related material, see
\cite{Downey:Fellows:2013}.
\emph{Bivincular patterns} generalize classical patterns even further
than vincular
patterns by adding a constraint on values.

We focus in this paper on pattern matching issues for
$(213,231)$-avoiding permutations that we call wedge permutations.
(\emph{i.e.}, those permutations that avoid both $213$ and $231$).
The number of $n$-wedge permutations is
$t_0 = 1$ for $n = 0$ and
$t_n =2^{n-1}$ for $n\geq 1$ \cite{Simion:Schmidt:EJC:1985},
as they are in bijection with binary word of size $n-1$.
On an individual basis,
the permutations that do not have an occurrence of the permutation pattern $231$
are exactly the \emph{stack-sortable permutations} and they are counted by
the Catalan numbers \cite{Knuth:1997:ACP:260999}.
%	A stack-sortable permutation is a permutation whose elements may be sorted by
%	an algorithm whose internal storage is limited to a single stack data structure.
%	As for $213$, it is well-known that
% 	if $\pi = \pi_1\pi\,\ldots\,\pi_n$ avoids $132$, then its complement
% 	$\pi' = (n+1-\pi_1)(n+1-\pi_2)\,\ldots\,(n+1-\pi_n)$ avoids $312$, and
% 	the reverse of $\pi'$ avoids $213$.
% From a combinatorial point of view,
% B\'ona \cite{Bona:ElJC:2012}
% showed the rather surprising fact that the cumulative number of
% occurrences of the classical patterns $231$ and $213$ are the same on the
% set of permutations avoiding $132$,
% beside the pattern based statistics $231$ and $213$
% do not have the same distribution on this set.
% \emph{Almost avoidance} has also been considered;
% a permutation $\pi$ almost avoids a set $X$ of permutations
% if there is a way to remove a single element of $\pi$ to get a permutation
% that avoids all elements in $P$
% and $L_n(P)$ denotes the set of permutations of length $n$ that almost avoid
% $P$.
% It is shown in \cite{Griffiths:Smith:Warren:PMA:2011} that
% $L_n(213, 231) = L_n(132, 231) = L_n(132, 312) = L_n(213, 312)$.
This paper is organized as follows.
In Section~\ref{section:Definitions} the needed definitions are presented.
Section~\ref{section:both are (213,231)-avoiding} is devoted to presenting
an online linear-time algorithm in case both 
permutations are wedge permutations,
whereas Section~\ref{section:sigma only avoid 231 and 213} focuses on the case
where only the pattern is a wedge permutation.
In Section~\ref{section:bivincular} we give a polynomial-time algorithm
for a bivincular wedge permutations pattern.
In Section~\ref{section:LCS} we consider the problem of finding the longest
wedge permutations pattern in permutations.

@@

A \textit{permutation} of size $n$ is a linear ordering of an ordered set of size $n$.
When writing the permutation we omit the $<$ sign, thus
writing the permutation as a word  :
$\pi = \pi_1\pi_2\,\ldots\,\pi_n$, whose elements are distinct
and usually consist of the integers $12\,\ldots\,n$. 
We let $\pi[i]$ stands for $\pi_i$.
Conveniently, we let
$\pi[i:j]$ stands for
$\pi_i\pi_{i+1}\,\ldots\,\pi_j$,
$\pi[:j]$ stands for $\pi[1:j]$ and
$\pi[i:]$ stands for $\pi[i:n]$.

We need to consider both the natural order of the set
and the linear order given by the permutation.
When talking about an element,
we refer to the relative position in
the natural order of the set
as its value 
and 
we refer to the relative position
in the order given by the permutation 
as its position.
We write 
$\pi[i]$ to indicate the value of the element 
at position $i$.  
For example, in the permutation $\pi = 51342$,
The element $1$ is at position $2$
and the element at position $1$ has for value $5$.

It is convenient to use a geometric representation of permutation
to ease the understanding of algorithms. 
The geometric representation corresponds to the set of points with coordinate $(i,\pi[i])$
(see Figure \ref{example: pattern matching}).

When an element $e_1$ is smaller (resp. bigger)
than an element $e_2$  by value (in the natural order of the set),
we say that $e_1$ is below (resp. above) $e_2$,
as a reference to the figure of the permutation.
Besides, the element with the biggest value is called
the topmost element and the element with the smallest value
is called the bottommost element.
For example, in the permutation $\pi = 51342$,
$2$ is below $4$, the topmost element is $5$
and the bottommost element is $1$.

When an element $e_1$ is smaller (resp. bigger)
than an element $e_2$  by position (in the order given by the permutation),
we say that $e_1$ is on the left of (resp. on the right of) $e_2$.
Besides, the element with the biggest position is called the rightmost element
and the element with the smallest position is called the leftmost element.
For example, in the permutation $\pi = 51342$,
$5$ is on the left of $3$, the leftmost element is $5$
and the rightmost element is $2$.

@@

A permutation $\sigma$ is said to \emph{occur} in the permutation $\pi$, written $\sigma \preceq \pi$,
if there exists a subsequence of (not necessarily consecutive)
elements of $\pi$ that has the same relative order as $\sigma$.
Otherwise, $\pi$ is said to \emph{avoid} the permutation $\sigma$.
For example, the permutation $\pi = 391867452$
has an occurrence of the pattern $\sigma = 51342$,
as can be seen in the highlighted subsequence of
$\pi = 3\mathbf{9}\mathbf{1}8\mathbf{6}\mathbf{7}\mathbf{4}52$
(or
$\pi = 3\mathbf{9}\mathbf{1}8\mathbf{6}\mathbf{7}4\mathbf{5}2$
or
$\pi = 3\mathbf{9}\mathbf{1}8\mathbf{6}\mathbf{7}45\textbf{2}$
or
$\pi = 3\mathbf{9}\mathbf{1}867\textbf{4}\textbf{5}\mathbf{2}$
).
Each subsequence $91674$,
$91675$,
$91672$,
$91452$,
 in $\pi$ is called an
\emph{occurrence}
of $\sigma$.
Since the permutation $\pi = 391867452$  contains no increasing subsequence of
length four, $\pi$ avoids $1234$.

For an occurrence $s$ of $\sigma$ in $\pi$, 
we say that $\pi[j]$ is a matching of $\sigma[i]$ or 
that $\sigma[i]$ is matched to $\pi[j]$  
if and only if $\pi[j]=s[i]$.

@@

An \emph{ascent element} of an $n$-permutation $\pi \in S_n$ is any element
$\pi[i]$ such that $i<n$ and $\pi[i] < \pi[i+1]$.
For example, the permutation
$3452167$ has ascents $3$, $4$, $1$ and $6$.
Similarly, a \emph{descent} is any element
$\pi[i]$ such that $i<n$ and $\pi[i] > \pi[i+1]$.
So for every $1 \leq i < n$, $\pi[i]$  is either an ascent or a descent of
$\pi$.
%To clarify the exposition,
%we let $\ustep$ and $\dstep$ denote an ascend and a descend, respectively.
%The \emph{stripe} $s_\pi$ of a permutation $\pi \in S_n$ is the word
%$\stripe{\pi}{1} \stripe{\pi}{2} \ldots \stripe{\pi}{n-1} \in \{\ustep,\dstep\}^{n-1}$
%defined by
%$\stripe{\pi}{i}= \ustep$ if $i$ an ascent in $\pi$ and
%$\stripe{\pi}{i} = \dstep$ if $i$ a descent in $\pi$.
%For example the stripe of the permutation
%$\pi = 981234765$
%is $\stripew{\pi} = \dstep\dstep\ustep\ustep\ustep\ustep\dstep\dstep$.

A $\textit{right to left maximum}$ (abbreviate RLMax) of $\pi$ is an element that does not have any element above it at its right (see Figure $\ref{fig:left to right maxima}$). Formally, $\pi[i]$ is a RLMax if and only if $\pi[i]$ is topmost element of $\pi[i:]$. Similarly $\pi[i]$ is a $\textit{left to right minima}$ (abbreviate RLMin) if and only if $\pi[i]$ is the bottommost element of $\pi[i:]$.

@@

	\caption[Example pattern matching]{
		The element is a RLMax if and only if the dashed area is empty.} 
	\label{fig:left to right maxima}
\end{figure}

A \emph{bivincular pattern} $\sigma$
of length $k$ is a permutation in $S_k$ written in
two-line notation
(the top row is $12\,\ldots\,k$ and the bottom row
is a permutation $\sigma_1\sigma_2\,\ldots\,\sigma_k$)
with overlined elements in the top row and underlined element in the bottom row.
We have the following conditions on the top and bottom rows
of $\sigma$, as see in \cite{Kitaev:book:2011} in Definition 1.4.1:
\begin{itemize}
	\item
	If the bottom line of $\sigma$ contains
	$\underline{\sigma_i\sigma_{i+1}\,\ldots\,\sigma_j}$
	then the elements corresponding to
	$\sigma_i\sigma_{i+1}\,\ldots\,\sigma_j$ in a occurrence of
	$\sigma$ in $\pi$ must be adjacent, whereas there is
	no adjacency condition for
	non-underlined consecutive elements.
	Moreover if the bottom row of $\sigma$ begins with
	$_\llcorner{\sigma_1}$ then any occurrence of $\sigma$
	in a permutation $\pi$ must begin with the leftmost
	element of $\pi$,
	and
	if the bottom row of $\sigma$ begins with
	${\sigma_k}_\lrcorner$ then any occurrence of $\sigma$
	in a permutation $\pi$ must end with the rightmost
	element of $\pi$.
	\item
	If the top line of $\sigma$ contains
	$\overline{i\,i+1\,\ldots\,j}$ then the elements corresponding to
	$i, i+1, \ldots, j$ in an
	occurrence of $\sigma$ in $\pi$ must be adjacent in values,
	whereas there is no value adjacency restriction for non-overlined
	elements.
	Moreover, if the top row of $\sigma$ begins with
	$^\ulcorner{1}$ then
	any occurrence of $\sigma$ is a permutation $\pi$ must contain 
	the bottommost element of $\pi$, and
	if top row of $\sigma$ ends with $k^\urcorner$ then
	any occurrence of $\sigma$ is a permutation $\pi$ must contain
	the topmost element of $\pi$.
\end{itemize}

For example,
let
$\sigma = \BV{1\overline{23}4\urcorner}{\llcorner 21\underline{43}  }$.
In $3217845$, $\textbf{32}17\textbf{84}5$ is an occurrence  of $\sigma$ but
 $3\textbf{21}\textbf{7}8\textbf{4}5$ is not.
The best general reference is \cite{Kitaev:book:2011}.

Geometrically, We represent underlined and overlined elements by \textit{forbidden areas}.
If two elements are overlined then we draw a vertical forbidden areas between those two points.
In an occurrence, we also draw the forbidden area between the matching of two overlined elements.
This area must be empty for an occurrence to be called so. 
If the area is empty then there is no element between them when reading the permutation
from bottom to top, 
in other words the matching elements are consecutive in value.
If two elements are underlined then we draw an horizontal forbidden area.
(See figure \ref{example:bivincular pattern matching}).

@@

    	\caption[Example pattern matching]{
    		From left to right,
    		the bivincular pattern $\sigma = \BV{1\overline{23}4\urcorner}{\llcorner 21\underline{43}  }$, An occurrence of $\sigma$ in $3216745$, An occurrence of $2143$ in $3216745$ but not an occurrence of $\sigma$ in $3216745$ because the point $(1,3)$ and $(5,7)$ are in the forbidden area.} 
    	\label{example:bivincular pattern matching}
\end{figure}

>>>>

\newcommand{\PM}{PM}

@@

%\DeclareMathOperator{\run}{run}
\DeclareMathOperator{\block}{block}
\DeclareMathOperator{\LMEi}{LMEp}

\DeclareMathOperator{\factor}{factor}

@@

%Pattern Marching
\DeclareMathOperator{\PMa}{PM}
%\newcommand{\PM}[6]{\PMa_{{#1}}^{{#2},{#3},{#4}}({#5},{#6})}

@@

\title{Permutation Pattern matching in\\ $(213,231)$-avoiding permutations}
\author{Both Emerite NEOU, Romeo RIZZI}
\date{}

\author{%
	Both Emerite Neou\thanks{On a Co-tutelle  Agreement with the Department of Mathematics of the University of Trento}\inst{1}
  Romeo Rizzi\inst{2} \and
  St\'ephane Vialette\inst{1}
}% end author
\institute{%
	Universit\'e Paris-Est, LIGM (UMR 8049), CNRS, UPEM, ESIEE Paris, ENPC,
	F-77454, Marne-la-Vallée, France\\
  \email{\{neou,vialette\}@univ-mlv.fr}
  \and
  Department of Computer Science,
  Universit\`a degli Studi di Verona, Italy \\
  \email{romeo.rizzi@univr.it}
}% end institute

@@

\begin{abstract}
Given permutations $\sigma$ of size $k$ and $\pi$ of size $n$ with $k<n$, the
\emph{permutation pattern matching} problem is to decide whether $\sigma$ occurs
$\pi$ as an order-isomorphic subsequence.
We give a linear-time algorithm in case both $\pi$ and $\sigma$ avoid
the two size-$3$ permutations $213$ and $231$.
For the special case where only $\sigma$ avoids $213$ and $231$, we present a
$O(max(kn^2,n^2\log(\log(n)))$-time algorithm. 
We extend our research to bivincular patterns that avoid $213$ and $231$ and present a $O(kn^4)$-time algorithm.
Finally we look at the related problem of the longest subsequence which avoids $213$ and $231$.  
\end{abstract}

@@

A permutation $\sigma$ is said to occur in another permutation $\pi$,
denoted $\sigma \preceq \pi$,
if there exists a subsequence of elements of $\pi$ that has the same relative
order as $\sigma$.
Otherwise, $\pi$ is said to \emph{avoid} the permutation $\sigma$.
For example the permutation $123$ (resp. $321$) occurs in a second permutation if
the second permutation has an increasing (resp. a decreasing) subsequence of size $3$.
Similarly, $213$ occurs in $6152347$, but $231$ does not occurs in $6152347$.
During the last decade, the study of patterns in permutations has
become a very active area of research \cite{Kitaev:book:2011} and
a whole annual conference (\textsc{Permutation Patterns}) focuses
on patterns in permutations.

We consider here the so-called \emph{permutation pattern matching} problem
(also sometimes referred to as the \emph{pattern involvement problem}):
Given two permutations $\sigma$ of size $k$ and $\pi$ of size $n$, the problem is to decide whether
$\sigma \preceq \pi$ (the problem is attributed to Wilf in \cite{Bose:Buss:Lubiw:1998}).
The permutation pattern matching is known to be \NPhard~\cite{Bose:Buss:Lubiw:1998}.
It is however polynomial-time solvable by brute-force enumeration
if $\sigma$ has bounded size.
Improvements to this algorithm were presented in
\cite{Albert:Aldred:Atkinson:Holton:ISAAC:2001} and
\cite{Ahal:Rabinovich:2008},
the latter describing a $O(n^{0.47k+o(k)})$-time algorithm.
Bruner and Lackner \cite{DBLP:journals/corr/abs-1204-5224}
gave a fixed-parameter algorithm solving the permutation pattern matching problem with
an exponential worst-case runtime of $O(1.52^{n}.n.k)$.
This is an improvement upon the $O(k {n \choose k})$ runtime required by
brute-force search without imposing restrictions on $\sigma$ and $\pi$,
where one has to enumerate all the ${n \choose k}$ different subsequences of size $k$ in $\pi$
and test if one of them has the same relative
order as $\sigma$.  
A recent major step was taken by Marx and Guillemot
\cite{Guillemot:Marx:SODA:2014}.
They showed that
the permutation pattern matching is fixed-parameter tractable (FPT) for by the size of $\sigma$.

A few particular cases of the permutation pattern matching problem have been attacked successfully.
The case of any increasing pattern is solvable in 
$O(n \log \log k)$ time in the RAM model \cite{Crochemore:Porat:2010},
improving the previous 30-year bound of $O(n \log k)$.
% (The algorithm also improves on the previous
% $O(|\pi| \log \log |\pi|)$ bound.)
Furthermore, the patterns $132$, $213$, $231$, $312$ can all be handled in linear-time
by stack sorting algorithms.
Any pattern of size $4$ can be detected in $O(n \log n)$ time
\cite{Albert:Aldred:Atkinson:Holton:ISAAC:2001}.
Algorithmic issues for $321$-avoiding patterns matching for permutations have been investigated in
\cite{Guillemot:Vialette:ISAAC:2009} and more recently in \cite{2015arXiv151006051A}.
The permutation pattern matching problem is also solvable in
polynomial-time for separable patterns \cite{Ibarra:1997,Bose:Buss:Lubiw:1998}
(see also \cite{Bouvel:Rossin:Vialette:CPM:2007} for the longest common subsequence problem and related ones on separable permutations).
Separable permutations are permutations that do not have an occurrence of
$2413$ nor $3142$, and they are enumerated by the Schr{\"o}der numbers.
%	(sequence A006318 in OEIS).
%	To the best of our knowledge,
%	separable permutations first arose in the work of
%	Avis and Newborn~\cite{Avis:Newborn:1981},
%	who showed that they are precisely the permutations which can be sorted by an
%	arbitrary number of pop-stacks in series,
%	where a pop-stack is a restricted form of stack in which any pop operation
%	pops all items at once.
(Notice that the separable permutations include as a special case the
stack-sortable permutations, which avoid the pattern $231$.)

There exist many generalisations of patterns that are worth considering
in the context of algorithmic issues in pattern matching
(see \cite{Kitaev:book:2011} for an up-to-date survey).
\emph{Vincular patterns}, also called
\emph{generalized patterns},
resemble (classical) patterns with the additional constraint that some of the elements in
a occurrence must be consecutive in positions.
Of particular importance in our context,
Bruner and Lackner \cite{DBLP:journals/corr/abs-1204-5224}
proved that deciding whether a vincular pattern
$\sigma$ of size $k$ occurs in a longer permutation
$\pi$ is $W[1]$-complete for
parameter $k$;
for an up-to-date survey of the $W[1]$ class and related material, see
\cite{Downey:Fellows:2013}.
\emph{Bivincular patterns} generalize classical patterns even further
than vincular
patterns by adding a constraint on values.

We focus in this paper on pattern matching issues for
$(213,231)$-avoiding permutations that we call wedge permutations.
(\emph{i.e.}, those permutations that avoid both $213$ and $231$).
%The number of $n$-wedge permutations is
%$t_0 = 1$ for $n = 0$ and
%$t_n =2^{n-1}$ for $n\geq 1$ \cite{Simion:Schmidt:EJC:1985},
%as they are in bijection with binary word of size $n-1$.
%On an individual basis,
%the permutations that do not have an occurrence of the permutation pattern $231$
%are exactly the \emph{stack-sortable permutations} and they are counted by
%the Catalan numbers \cite{Knuth:1997:ACP:260999}.
%	A stack-sortable permutation is a permutation whose elements may be sorted by
%	an algorithm whose internal storage is limited to a single stack data structure.
%	As for $213$, it is well-known that
% 	if $\pi = \pi_1\pi\,\ldots\,\pi_n$ avoids $132$, then its complement
% 	$\pi' = (n+1-\pi_1)(n+1-\pi_2)\,\ldots\,(n+1-\pi_n)$ avoids $312$, and
% 	the reverse of $\pi'$ avoids $213$.
% From a combinatorial point of view,
% B\'ona \cite{Bona:ElJC:2012}
% showed the rather surprising fact that the cumulative number of
% occurrences of the classical patterns $231$ and $213$ are the same on the
% set of permutations avoiding $132$,
% beside the pattern based statistics $231$ and $213$
% do not have the same distribution on this set.
% \emph{Almost avoidance} has also been considered;
% a permutation $\pi$ almost avoids a set $X$ of permutations
% if there is a way to remove a single element of $\pi$ to get a permutation
% that avoids all elements in $P$
% and $L_n(P)$ denotes the set of permutations of length $n$ that almost avoid
% $P$.
% It is shown in \cite{Griffiths:Smith:Warren:PMA:2011} that
% $L_n(213, 231) = L_n(132, 231) = L_n(132, 312) = L_n(213, 312)$.
This paper is organized as follows.
In Section~\ref{section:Definitions} the needed definitions are presented.
Section~\ref{section:both are (213,231)-avoiding} is devoted to presenting
an online linear-time algorithm in case both 
permutations are wedge permutations,
whereas Section~\ref{section:sigma only avoid 231 and 213} focuses on the case
where only the pattern is a wedge permutation.
In Section~\ref{section:bivincular} we give a polynomial-time algorithm
for a bivincular wedge permutation pattern.
In Section~\ref{section:LCS} we consider the problem of finding the longest
wedge permutation pattern in permutations.

@@

A \textit{permutation} of size $n$ is a linear ordering of an ordered set of size $n$.
When writing the permutation we omit the $<$ sign, thus
writing the permutation as a word:
$\pi = \pi_1\pi_2\,\ldots\,\pi_n$, whose elements are distinct
and usually consist of the integers $1,2,\ldots,n$.

We need to consider both the natural order of the set
and the linear order given by the permutation.
When talking about an element,
we refer to the relative position in
the natural order of the set
as its value 
and 
we refer to the relative position
in the order given by the permutation 
as its position.
We write 
$\pi[i]$ to indicate the value of the element 
at position $i$ (\emph{i.e.} $\pi_i$).  
For example, in the permutation $\pi = 51342$,
The element $1$ is at position $2$
and the element at position $1$ has for value $5$.
Conveniently, we let
$\pi[i:j]$ stand for
$\pi_i\pi_{i+1}\,\ldots\,\pi_j$,
$\pi[:j]$ stand for $\pi[1:j]$ and
$\pi[i:]$ stand for $\pi[i:n]$.

It is convenient to use a geometric representation of permutations
to ease the understanding of algorithms. 
The geometric representation corresponds to the set of points with coordinate $(i,\pi[i])$
(see Figure \ref{example: pattern matching}).

When an element $\pi[\alpha]$ is smaller (resp. larger)
than an element $\pi[\beta]$  by value (in the natural order of the set),
we say that $\pi[\alpha]$ is below (resp. above) $\pi[\beta]$,
as a reference to the geometric representation of the permutation.
Besides, the element with the largest value is called
the topmost element and the element with the smallest value
is called the bottommost element.
For example, in the permutation $\pi = 51342$,
$2$ is below $4$, the topmost element is $5$
and the bottommost element is $1$.

When an element $\pi[\alpha]$ is smaller (resp. larger)
than an element $\pi[\beta]$  by position (in the order given by the permutation),
we say that $\pi_\alpha$ is on the left of (resp. on the right of) $\pi[\beta]$.
Besides, the element with the largest position is called the rightmost element
and the element with the smallest position is called the leftmost element.
For example, in the permutation $\pi = 51342$,
$5$ is on the left of $3$, the leftmost element is $5$
and the rightmost element is $2$.

@@

A permutation $\sigma$ is said to \emph{occur} in the permutation $\pi$, written $\sigma \preceq \pi$,
if there exists a subsequence of (not necessarily consecutive)
elements of $\pi$ that has the same relative order as $\sigma$.
Such subsequence is called an \emph{occurrence} of $\sigma$ in $\pi$.
Otherwise, $\pi$ is said to \emph{avoid} the permutation $\sigma$.
For example, the permutation $\pi = 391867452$
has an occurrence of the pattern $\sigma = 51342$,
as can be seen in the highlighted subsequence of
$\pi = 3\mathbf{9}\mathbf{1}8\mathbf{6}\mathbf{7}\mathbf{4}52$
(or
$\pi = 3\mathbf{9}\mathbf{1}8\mathbf{6}\mathbf{7}4\mathbf{5}2$
or
$\pi = 3\mathbf{9}\mathbf{1}8\mathbf{6}\mathbf{7}45\textbf{2}$
or
$\pi = 3\mathbf{9}\mathbf{1}867\textbf{4}\textbf{5}\mathbf{2}$
).
Since the permutation $\pi = 391867452$  contains no increasing subsequence of
size four, $\pi$ avoids $1234$.

For an occurrence $s$ of $\sigma$ in $\pi$, 
we say that $\pi[j]$ correspond $\sigma[i]$ in the occurrence or 
that $\sigma[i]$ is matched to $\pi[j]$  in the occurrence
if and only if $\pi[j]=s[i]$.

@@

%An \emph{ascent element} of an $n$-permutation $\pi \in S_n$ is any element
%$\pi[i]$ such that $i<n$ and $\pi[i] < \pi[i+1]$.
%For example, the permutation
%$3452167$ has ascents $3$, $4$, $1$ and $6$.
%Similarly, a \emph{descent} is any element
%$\pi[i]$ such that $i<n$ and $\pi[i] > \pi[i+1]$.
%So for every $1 \leq i < n$, $\pi[i]$  is either an ascent or a descent of
%$\pi$.

%To clarify the exposition,
%we let $\ustep$ and $\dstep$ denote an ascend and a descend, resp..
%The \emph{stripe} $s_\pi$ of a permutation $\pi \in S_n$ is the word
%$\stripe{\pi}{1} \stripe{\pi}{2} \ldots \stripe{\pi}{n-1} \in \{\ustep,\dstep\}^{n-1}$
%defined by
%$\stripe{\pi}{i}= \ustep$ if $i$ an ascent in $\pi$ and
%$\stripe{\pi}{i} = \dstep$ if $i$ a descent in $\pi$.
%For example the stripe of the permutation
%$\pi = 981234765$
%is $\stripew{\pi} = \dstep\dstep\ustep\ustep\ustep\ustep\dstep\dstep$.

A $\textit{right-to-left maximum}$ (abbreviated to RLMax) of $\pi$ is an element that does not have any element above it and to its right (see Figure $\ref{fig:left to right maxima}$). Formally, $\pi[i]$ is a RLMax if and only if $\pi[i]$ is topmost element of $\pi[i:]$. Similarly $\pi[i]$ is a $\textit{right-to-left minimum}$ (abbreviate RLMin) if and only if $\pi[i]$ is the bottommost element of $\pi[i:]$.

@@

	\caption[Example pattern matching]{
		The element is a RLMax if and only if the hashed area is empty.} 
	\label{fig:left to right maxima}
\end{figure}

A \emph{bivincular permutation pattern} $\widetilde{\sigma}$
of size $k$ is a permutation written in
two-line notation
(the top row is $12\,\ldots\,k$ and the bottom row
is a permutation $\sigma_1\sigma_2\,\ldots\,\sigma_k$)
with overlined elements in the top row and underlined elements in the bottom row.
We have the following conditions on the top and bottom rows
of $\sigma$, as seen in \cite{Kitaev:book:2011} in Definition 1.4.1:
\begin{itemize}
	\item
	If the bottom line of $\sigma$ contains
	$\underline{\sigma_i\sigma_{i+1}\,\ldots\,\sigma_j}$
	then the elements corresponding to
	$\sigma_i\sigma_{i+1}\,\ldots\,\sigma_j$ in a occurrence of
	$\sigma$ in $\pi$ must be adjacent, whereas there is
	no adjacency condition for
	non-underlined consecutive elements.
	Moreover if the bottom row of $\sigma$ begins with
	$_\llcorner{\sigma_1}$ then any occurrence of $\sigma$
	in a permutation $\pi$ must begin with the leftmost
	element of $\pi$,
	and
	if the bottom row of $\sigma$ ends with
	${\sigma_k}_\lrcorner$ then any occurrence of $\sigma$
	in a permutation $\pi$ must end with the rightmost
	element of $\pi$.
	\item
	If the top line of $\sigma$ contains
	$\overline{i\,i+1\,\ldots\,j}$ then the elements corresponding to
	$i, i+1, \ldots, j$ in an
	occurrence of $\sigma$ in $\pi$ must be adjacent in values,
	whereas there is no value adjacency restriction for non-overlined
	elements.
	Moreover, if the top row of $\sigma$ begins with
	$^\ulcorner{1}$ then
	any occurrence of $\sigma$ in a permutation $\pi$ must contain 
	the bottommost element of $\pi$, and
	if top row of $\sigma$ ends with $k^\urcorner$ then
	any occurrence of $\sigma$ in a permutation $\pi$ must contain
	the topmost element of $\pi$.
\end{itemize}

Given a bivincular permutation pattern $\widetilde{\sigma}$,
$\sigma$ refers to the bottom line of $\widetilde{\sigma}$
without any element underlined.
For example,
let
$\widetilde{\sigma} = \BV{1\overline{23}4\urcorner}{\llcorner 21\underline{43}}$,
in $3217845$, $\textbf{32}17\textbf{84}5$ is an occurrence  of $\widetilde{\sigma}$ but $3\textbf{21}\textbf{7}8\textbf{4}5$ is not 
and $\sigma = 2143$.
The best general reference for bivincular pattern is \cite{Kitaev:book:2011}.

Geometrically, we represent underlined and overlined elements by \textit{forbidden areas}.
If two elements are overlined then we draw a horizontal forbidden area between those two points.
In an occurrence, we also draw the forbidden area between the matching of two overlined elements.
This area must be empty to have a valid occurrence of the bivincular pattern. 
If the area is empty then there is no element between them when reading the permutation
from bottom to top, 
in other words the matching elements are consecutive in value.
If two elements are underlined then we draw an vertical forbidden area.
(See Figure \ref{example:bivincular pattern matching}).

@@

    	\caption[Example pattern matching]{
    		From left to right,
    		the bivincular pattern $\widetilde{\sigma} = \BV{1\overline{23}4\urcorner}{\llcorner 21\underline{43}  }$, an occurrence of $\widetilde{\sigma}$ in $3216745$, an occurrence of $\sigma$ in $3216745$ but not an occurrence of $\widetilde{\sigma}$ because the points $(1,3)$ and $(5,7)$ are in the forbidden areas.} 
    	\label{example:bivincular pattern matching}
\end{figure}
